Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-09-26 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Revisions Ch1 through Ch5
Added photo for Investigator Challenge

In 2003, a list of zone and computational fluid dynamics (CFD) fire models was compiled by Olenick and Carpenter~\cite{Olenick:2003}. A similar survey was conducted in 1992, and Olenick has updated the list in 2014. There are 53 zone models listed, but only three are actively supported or maintained.  The survey also includes 22 field models of which 10 are listed as being actively supported.  All of the correlations and models have been to some extent been validated or checked against experimental data for a limited set of scenarios~\cite{ASTM_E1355}.

@@

Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.  A key limitation is the assumption by many of the researchers, or text authors, that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin, some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

@@

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 – 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained~\cite{Beyler:2009}.

@@

For the moment lets step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 in the morning on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older townhomes.  The socio-economics of the neighborhood are such the the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene, ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.''  You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his work place.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  The is some thermal damage and smoke deposits on the from of the siding on the porch.  The outer pane of the living room window is for the most part still intact. you notice that three of the four windows on the second floor have been removed.  On ``side D'' of the townhome there are no openings and no smoke or fire damage is present.  On the rear or ``side C'' of the house the door and windows are intact but open.  There is no smoke stains or thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is not repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There are a clear horizontal line of demarcation the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room was a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is not thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs the less damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focusing on the area with the most damage, the area with the sofa and chair.  As you are looking toward the ``B wall'' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  The corner formed by the wall behind the chair and the wall of the stairway has a ``V pattern'' burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The burned wire from the heater extends toward the wall outlet behind the sofa.

You have now developed your first hypothesis about the origin of the fire.  A radiant style, electric heater was turned on and left until it ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine if the heater switch was in the ``on'' position.  Now you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the right side back cushion of the sofa. The damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee table looks like a arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the sofa frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further is appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair clearly has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not be moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews what else can you do to eliminate the inconsistencies?   We will reexamine the in Chapter XX.

@@

A schematic drawing of the instrumentation arrangement is shown in Fig.~\ref{Drawing_Fire_Flamer}.  The thermocouples were mounted on a steel frame.  The frame enabled the thermocouples to be installed at 10~cm intervals from 10~cm to 120~cm above the burner/pan lip/fuel surface. At each 10~cm interval, there were five thermocouples: one centered above the burner/fuel surface and one centered above each edge of the burner/fuel surface, for a total of 70 thermocouples.  A plumb bob and steel tape measure, with a resolution of $\pm$~0.5~mm, were used to position and align the thermocouples.   The expanded uncertainty for the location of each thermocouple tip is estimated to be $\pm$~2.5~\%.  The same applies to the location of the heat flux sensor surfaces.

@@

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 100~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11~\% It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a ``square-wave''.

@@

Figure~\ref{Foam_Contours} shows two different contours for the polyurethane foam fueled experiment.  The results from polyurethane fueled fire characterization experiments have an increased amount of variability due to the nature of the ignition source and the fuel itself.  The polyurethane foam was readily ignited with a small pilot flame similar to that used to ignite the natural gas and the gasoline.  By comparison, a small flame ignited the gases flowing out the top of the natural gas burner and the flames from the natural gas cover entire burner surface within approximately one second.  A similar ignition sequence occurs with the gasoline pan fires. The polyurethane foam has a more complicated flame spread sequence that resulted in significant changes to the peak heat release rate and the time to peak heat release.  Once the polyurethane foam was ignited, it took approximately 60~s for the flame to spread across the top surface of the fuel.  As the flames spread horizontally across the fuel surface, the fire also burns down into the fuel.  In some cases the fire burned completely through the fuel in the center of the foam sample, resulting in a hollow flame at the time of the peak heat release rate.  An example of a contour plot, exhibiting the hollow or bifurcated flames is shown at right in Figure~\ref{Foam_Contours}. Refer back to Figure~\ref{flame_height_comp_FHPUF_1} to see photographic views of the bifurcated flames.  It seems that the movement of the flames as they flowed away from the fuel surface blurred out the bifurcation over time with the exception of the lower elevation areas that presented at the 90 and 100~\% contours.  Other methods of ignition for the polyurethane foam were considered in order to have a more uniform ignition of the top surface of the foam.  Each method would have resulted in a larger release of energy from the ignition source or a starter fuel.   Based on a few scoping experiments these methods did not improve the repeatability.

Based on the field data developed using Streams average flame heights for each fuel were determined.  In Table~\ref{tab:Average_Flame_Heights}, each of the values represents the average flame heights, given in mm with 95~\% confidence limits based on a Type A statistical analysis of the data, for each of the source fuels shown as the percentage of time the flame was visible at that given height.  For the natural gas fires, the averages are from ten 30~s segments, while the fire was monitored at approximately 80~kW.  In some cases more than one 30~s segment was sampled per heat release rate experiment.  Sixteen 30~s video segments, which included the peak heat release rate, were successfully analyzed from the seventeen gasoline heat release rate experiments.  One of the videos was unusable for analysis due to reflected light which could not be edited out of the images.  For same reason, the polyurethane foam flame height averages are the result from eight of the ten heat release rate experiments.

@@

\begin{table}
\small
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Measurement     &   \multicolumn{3}{|c|}{Fuels} \\ \cline{2-4}
type                    &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Avg. Peak HRR (kW)      &       80 $\pm$ 9 (11~\%)   &	80 $\pm$ 24 (30~\%) 	& 45 $\pm$ 21 (47~\%)   \\
Total HR (MJ)           &   	23.3 $\pm$ 1.6 (7~\%)   &	13.6 $\pm$ 0.6 (4~\%) 	& 4.1 $\pm$ 0.2 (5~\%)   \\
Avg Peak HF (kW/m$^2$)   &   	2.0 $\pm$ 0.5 (25~\%)   &	3.6 $\pm$ 0.9 (25~\%) 	& 2.5 $\pm$ 0.5 (20~\%)   \\
Median FH (m)           &   	705 $\pm$ 106 (15~\%)   &	695 $\pm$ 95 (14~\%) 	& 470 $\pm$ 160 (34~\%)   \\
Avg Peak Temp. (C)      &   	410 $\pm$ 103 (25~\%)   &	335 $\pm$ 41 (12~\%) 	& 334 $\pm$ 106 (32~\%)   \\
\hline
\end{tabular}
 \caption[Select measurements for the three source fires]{Select measurements for the three source fires. Each measurement is shown with 95~\% confidence limits both in terms of a value and a percentage.}
 \label{tab:Source_fire_summary}
\end{table}

@@

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array,a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

The results are presented for each of the four burner positions, (0.61~m, 0.31~m, 0.15~m and 0.0~m), and each of the three fuels in Appendix B.  For the natural gas fueled experiments only two replicates were conducted at each position given the higher level of control and repeatability.  For the gasoline and polyurethane fueled experiments, three replicates were conducted at each position. The position is measured as the distance between the edge of the burner closest to the wall and the wall.  The results of the wall temperatures and heat fluxes are presented as averages of the replicate experiments.  In the sections below, a full set of the averaged data from the natural gas fueled experiments will be presented to demonstrate the impact on distance from the wall on the temperatures in the plume as well as the conditions measured on the wall. Then composite graphs of the wall temperatures and heat flux measure on the wall versus distance will be shown for each fuel type.

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs after approximately 3~min after ignition.  A review of the heat release rates and the test videos do not yield any clues as to the reason for this decrease which occurred in both experiments.  In both experiments the heat release rate was steady and the video did not show any change in flame movement.
Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100~$^\circ$C.

@@

Figure~\ref{NCTW_Surf_Edge_comp} presents the temperature of the vertical array of thermocouples positioned on the wall along the edge of the burner for each of the three fuels.  The data is treated in the same manner as in Figure~\ref{NCTW_Surf_Center_comp}.  For each fuel the average peak temperatures at the burner edge are significantly lower when the centerline temperatures when the burner is positioned against the wall.  As the burners were moved away from the wall the temperatures at the centerline and the edge thermocouple arrays begin to converge.

@@

Figures~\ref{NCTW_HF_Center_Comp} and \ref{NCTW_HF_Edge_Comp} show the results from the arrays of total heat flux gauges located along the vertical centerline and the edge of the burners for each fuel.  The sensing surface of the gauges were installed flush with the wall surface and exposed to the fire. In each graph, the averaged peak heat flux for each of the replicate experiments and each position is shown.  Moving from 0.2~m above the top surface of the burner up to 1.2~m above the top of the burner surface in increments of 0.2~m.  Reviewing the graph for each fuel demonstrates how the heat flux at the wall decreased as the position of the burner was moved away from the wall.  (Review video and tie in ST TC graph with edge peak NG at 0.15)

@@

The complete set of results from the instrumented gypsum wallboard experiments are provided in Appendix C. Table~\ref{tab:IWGB_experiments} lists the experiments that were conducted for each fuel type.  In this chapter, as in the previous chapter the averaged peak values for temperatures and heat fluxes will presented at varying distances from the wall.  For the gasoline and polyurethane fires, an additional position, 0.08~m from the wall was added.

Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set} and \ref{IWGB_NG_TC_Backside_set} show the values for the natural gas burner experiments with the gypsum wallboard.

>>>>

In 2003, a list of zone and computational fluid dynamics (CFD) fire models was compiled by Olenick and Carpenter~\cite{Olenick:2003}. A similar survey was conducted in 1992, and Olenick updated the list in 2014. There are 53 zone models listed, but only three are actively supported or maintained.  The survey also includes 22 field models of which 10 are listed as being actively supported.  All of the correlations and models have, to some extent, been validated or checked against experimental data for a limited set of scenarios~\cite{ASTM_E1355}.

@@

Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount of research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.  A key limitation is the assumption by many of the researchers, or text authors, that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin, some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

@@

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 {\textendash} 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained~\cite{Beyler:2009}.

@@

\section{The Fire Investigators' Challenge}

For the moment let's step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 in the morning on a cool fall morning and you are being called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older townhomes.  The socio-economics of the neighborhood are such the the people that live there have many life struggles. Enroute you learn that the fire department rescued a mother and child from the home and they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted for at this time.  Upon arrival you have a conversation with the police and fire officers on the scene.  You are provided with information regarding what the first firefighters saw on the scene; ``all of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.'' You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know that the address is familiar to them for domestic disturbance calls for the husband and wife fighting but no charges have been filed.  They are looking for the husband and have sent a patrol car to his place of work.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On ``side D'' of the townhome there are no openings and no smoke or fire damage is present.  On the rear or ``side C'' of the house the door and windows are intact but open.  There are no smoke stains or thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure, the fire department is repacking their equipment and are preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There are is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the townhome and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the ``D wall'' of the living room there is a wood framed sofa.  The cushions have thermal damage, from pyrolysis.  It is not clear that they were on fire as there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs the less thermal damage you find.

You head back down the stairs to further examine the living room which appears to be the room of fire origin. Your are focusing on the area with the most damage, the area with the sofa and chair.  As you are looking toward the ``B wall'' of the living room you note that the right side of the chair is badly burned and it is burned from the floor level up.  The same is true from the left side of the sofa.  In fact, the left side of the sofa has more fire damage to the wood frame of the sofa than the right side of the sofa frame.  The corner formed by the wall behind the chair and the wall of the stairway has a ``V pattern'' burned into it, from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The burned wire from the heater extends toward the wall outlet behind the sofa. Figure~\ref{EXT_LRDetails} shows some of these details.

You have now developed your first hypothesis about the origin of the fire.  A radiant style, electric heater was turned on.  It was too close to the furniture and ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine if the heater switch was in the ``on'' position.  Now you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

There are different damage patterns behind the sofa.  On the right side there is fire damage on the wall behind the right side back cushion of the sofa. The damaged area is distinct from the other fire damage that exists on the wall behind the sofa. There are pieces of charred newspaper near the right arm of the sofa.  The burn mark on the top of the wood coffee table looks like an arrow that is pointing toward the corner.  The char on the legs of the coffee table is very directional and consistent with the heat coming from the corner.  Looking at the inside area of the sofa, you see that there is charred wood on the inside surface of the sofa frame while the outer portions of the wood frame are not charred.  This would indicate that the fire was inside the sofa at some point.  Further is appears that all of the foam cushioning is burnt off of the right side of the sofa, while there is a piece of polyurethane foam cushion on the left side of the sofa.  However that foam could have gotten moved during the fire suppression.

The chair has more unburned foam cushioning then the sofa does and the seat cushion is still in place with it's distinctive shape, so you know that cushion has not been moved.  The cushion was heated from the top down.  Perhaps the construction of the chair slowed the flame spread through the chair from the bottom up, if the heater was the ignition source.  Or the fire started in the sofa and the radiation from the flames and hot gas layer spread the fire to the chair  The heater still looks good as a source however now you have some fire damage that may not be consistent with the heater as the origin.  Besides interviews what else can you do to eliminate the inconsistencies?   The fire investigators' challenge we be reexamined in Chapter IX.

\begin{figure}[ht!]
	\centering
		\includegraphics[width=2.5in]{../Figures/EXT_SideA}
		\includegraphics[width=2.5in]{../Figures/INT_SideD}\\
		\hspace{3 mm}  \\
		\includegraphics[width=2.5in]{../Figures/INT_SideC}
		\includegraphics[width=2.5in]{../Figures/INT_SideA}\\
	\caption[Photographs of the exterior and interior of a townhouse post fire]{Photographs of the exterior and interior of a townhouse post fire. The three interior photos show the right side wall of the living room (Side D), the back wall of the living room (Side C), and the front wall of the living room (Side A).}
	\label{EXT_INTLR}
\end{figure}

\begin{figure}[ht!]
	\centering
		\includegraphics[width=2.5in]{../Figures/V_pattern}
		\includegraphics[width=2.5in]{../Figures/Heater}\\
		\hspace{3 mm}  \\
		\includegraphics[width=2.5in]{../Figures/Sofa}
		\includegraphics[width=2.5in]{../Figures/Table}\\
	\caption[Photographs of the some of the details of the living room fire area]{Photographs of the some of the details of the living room fire area including the V pattern, heater behind the chair and sofa, the varied thermal damage on the sofa, and the thermal damage to the coffee table and the interior of the sofa frame.}
	\label{EXT_LRDetails}
\end{figure}

@@

A schematic drawing of the instrumentation arrangement is shown in Figure~\ref{Drawing_Fire_Flamer}.  The thermocouples were mounted on a steel frame.  The frame enabled the thermocouples to be installed at 10~cm intervals from 10~cm to 120~cm above the burner/pan lip/fuel surface. At each 10~cm interval, there were five thermocouples: one centered above the burner/fuel surface and one centered above each edge of the burner/fuel surface, for a total of 70 thermocouples.  A plumb bob and steel tape measure, with a resolution of $\pm$~0.5~mm, were used to position and align the thermocouples.   The expanded uncertainty for the location of each thermocouple tip is estimated to be $\pm$~2.5~\%.  The same applies to the location of the heat flux sensor surfaces.

@@

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 300~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11~\% It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a ``square-wave''.

@@

Figure~\ref{Foam_Contours} shows two different contours for the polyurethane foam fueled experiment.  The results from polyurethane fueled fire characterization experiments have an increased amount of variability due to the nature of the ignition source and the fuel itself.  The polyurethane foam was readily ignited with a small pilot flame similar to that used to ignite the natural gas and the gasoline.  By comparison, a small flame ignited the gases flowing out of the top of the natural gas burner and the flames from the natural gas cover the entire burner surface within approximately one second.  A similar ignition sequence occurs with the gasoline pan fires. The polyurethane foam has a more complicated flame spread sequence that resulted in significant changes to the peak heat release rate and the time to peak heat release.  Once the polyurethane foam was ignited, it took approximately 60~s for the flame to spread across the top surface of the fuel.  As the flames spread horizontally across the fuel surface, the fire also burns down into the fuel.  In some cases the fire burned completely through the fuel in the center of the foam sample, resulting in a hollow flame at the time of the peak heat release rate.  An example of a contour plot, exhibiting the hollow or bifurcated flames is shown at right in Figure~\ref{Foam_Contours}. Refer back to Figure~\ref{flame_height_comp_FHPUF_1} to see photographic views of the bifurcated flames.  It seems that the movement of the flames as they flowed away from the fuel surface blurred out the bifurcation over time with the exception of the lower elevation areas that presented at the 90 and 100~\% contours.  Other methods of ignition for the polyurethane foam were considered in order to have a more uniform ignition of the top surface of the foam.  Each method would have resulted in a larger release of energy from the ignition source or a starter fuel.   Based on a few scoping experiments these methods did not improve the repeatability.

Based on the field data developed using Streams average flame heights for each fuel were determined.  In Table~\ref{tab:Average_Flame_Heights}, each of the values represents the average flame heights, given in mm with 95~\% confidence limits based on a Type A statistical analysis of the data, for each of the source fuels shown as the percentage of time the flame was visible at that given height.  For the natural gas fires, the averages are from ten 30~s segments, while the fire was monitored at approximately 80~kW.  In some cases more than one 30~s segment was sampled per heat release rate experiment.  Sixteen 30~s video segments, which included the peak heat release rate, were successfully analyzed from the seventeen gasoline heat release rate experiments.  One of the videos was unusable for analysis due to reflected light which could not be edited out of the images.  For the same reason, the polyurethane foam flame height averages are the result from eight of the ten heat release rate experiments.

@@

\begin{table}
\small
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Measurement     &   \multicolumn{3}{|c|}{Fuels} \\ \cline{2-4}
type                    &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Avg. Peak HRR (kW)      &       80 $\pm$ 9 (11~\%)   &	80 $\pm$ 24 (30~\%) 	& 45 $\pm$ 21 (47~\%)   \\
Total HR (MJ)           &   	23.3 $\pm$ 1.6 (7~\%)   &	13.6 $\pm$ 0.6 (4~\%) 	& 4.1 $\pm$ 0.2 (5~\%)   \\
Avg Peak HF (kW/m$^2$)   &   	2.0 $\pm$ 0.5 (25~\%)   &	3.6 $\pm$ 0.9 (25~\%) 	& 2.5 $\pm$ 0.5 (20~\%)   \\
Median FH (mm)           &   	705 $\pm$ 106 (15~\%)   &	695 $\pm$ 95 (14~\%) 	& 470 $\pm$ 160 (34~\%)   \\
Avg Peak Temp. (C)      &   	410 $\pm$ 103 (25~\%)   &	335 $\pm$ 41 (12~\%) 	& 334 $\pm$ 106 (32~\%)   \\
\hline
\end{tabular}
 \caption[Select measurements for the three source fires]{Select measurements for the three source fires. Each measurement is shown with 95~\% confidence limits both in terms of a value and a percentage.}
 \label{tab:Source_fire_summary}
\end{table}

@@

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array, a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

The results are presented for each of the four burner positions, (0.61~m, 0.31~m, 0.15~m and 0.0~m), and each of the three fuels.  For the natural gas fueled experiments only two replicates were conducted at each position given the higher level of control and repeatability.  For the gasoline and polyurethane fueled experiments, three replicates were conducted at each position. The position is measured as the distance between the edge of the burner closest to the wall and the wall.  The results of the wall temperatures and heat fluxes are presented as averages of the replicate experiments.  In the sections below, a full set of the averaged data from the natural gas fueled experiments will be presented to demonstrate the impact on distance from the wall on the temperatures in the plume as well as the conditions measured on the wall. Then composite graphs of the wall temperatures and heat flux measure on the wall versus distance will be shown for each fuel type.

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhpas as a result of increased convection on the face of the wall.

Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100~$^\circ$C.

@@

Figure~\ref{NCTW_Surf_Edge_comp} presents the temperature of the vertical array of thermocouples positioned on the wall along the edge of the burner for each of the three fuels.  The data is treated in the same manner as in Figure~\ref{NCTW_Surf_Center_comp}.  For each fuel the average peak temperatures at the burner edge are significantly lower than the centerline temperatures when the burner is positioned against the wall.  As the burners were moved away from the wall the temperatures at the centerline and the edge thermocouple arrays begin to converge.

@@

Figures~\ref{NCTW_HF_Center_Comp} and \ref{NCTW_HF_Edge_Comp} show the results from the arrays of total heat flux gauges located along the vertical centerline and the edge of the burners for each fuel.  The sensing surface of the gauges were installed flush with the wall surface and exposed to the fire. In each graph, the averaged peak heat flux for each of the replicate experiments and each position is shown.  Moving from 0.2~m above the top surface of the burner up to 1.2~m above the top of the burner surface in increments of 0.2~m.  Reviewing the graph for each fuel demonstrates how the heat flux at the wall decreased as the position of the burner was moved away from the wall.

@@

Table~\ref{tab:IWGB_experiments} lists the experiments that were conducted for each fuel type.  In this chapter, as in the previous chapter the averaged peak values for temperatures and heat fluxes will presented at varying distances from the wall.  For the gasoline and polyurethane fires, an additional position, 0.08~m from the wall was added.

A complete set of data temperature and heat flux data for the six natural gas burner experiments are provided in Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set},\ref{IWGB_NG_TC_Backside_set},\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 200~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed.  However at the point there was no opprotunity to rerun the experiment.
